\documentclass[12pt]{book}

\usepackage{mathtools}
\usepackage{quiver}
% \usepackage{algorithm}
% \usepackage{algpseudocode}
% \algrenewcommand\algorithmicrequire{\textbf{Input:}}
% \algrenewcommand\algorithmicensure{\textbf{Output:}}
% \algnewcommand\Not{\textbf{not} }
% \newcommand{\algorithmautorefname}{Algorithm} % autoref

% % http://paultaylor.eu/diagrams/
% % https://www.jmilne.org/not/Mdiagrams.pdf
\usepackage[small,nohug,heads=vee]{diagrams}
\diagramstyle[labelstyle=\scriptstyle]

\usepackage[inline]{enumitem}
\makeatletter
% This command ignores the optional argument for itemize and enumerate lists
\newcommand{\inlineitem}[1][]{%
\ifnum\enit@type=\tw@
    {\descriptionlabel{#1}}
  \hspace{\labelsep}%
\else
  \ifnum\enit@type=\z@
       \refstepcounter{\@listctr}\fi
    \quad\@itemlabel\hspace{\labelsep}%
\fi}
\makeatother

\usepackage{adjustbox}
\usepackage[headings]{fullpage}
\usepackage{xargs}
\usepackage{xspace}
\usepackage[doublespacing]{setspace}

\usepackage{hyperref}
\def\chapterautorefname{Chapter}
\def\sectionautorefname{Section}
\def\subsectionautorefname{Section}
\def\subsubsectionautorefname{Section}

\usepackage{microtype}
\usepackage[T1]{fontenc}

\usepackage{wrapfig}
\usepackage[font={sf}]{caption}
\usepackage[subrefformat=parens]{subcaption}

\usepackage{multirow}

\usepackage{xcolor}
\usepackage{graphicx}
\graphicspath{{_build/}{figures/}}

\usepackage{amsmath}
\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{defn}{Definition}[chapter]
\newtheorem{thm}{Theorem}[chapter]
% \newtheorem{proof}{Proof}[chapter]
\newtheorem{prop}{Proposition}[chapter]
\newtheorem{lmm}{Lemma}[chapter]
\newtheorem{ex}{Example}[chapter]
\newtheorem{fact}{Example}[chapter]
\newtheorem{corollary}{Corollary}[chapter]

\setlength{\marginparwidth}{2cm}
\usepackage[textsize=tiny]{todonotes}
\newcommand\remy[2][]{\todo[color=orange, #1]{\sffamily #2}}

\usepackage{libertine}
\usepackage[libertine]{newtxmath}
\usepackage[scaled=.85]{DejaVuSansMono}

\usepackage{listings}
\lstset{
  columns=fullflexible,
  keepspaces=true,
  showstringspaces=false,
  stringstyle=\slshape\color{green!40!black},
  basicstyle=\ttfamily\small,
  language=Python,
  morekeywords={*, self},
  deletekeywords=[2]{map, iter, vars, tuple, len},
  commentstyle=\slshape\color{black!60},
  tabsize=2,
  mathescape=true,
}

\lstdefinelanguage{Rust}{
  sensitive,
  morecomment=[l]{//},
  morecomment=[s]{/*}{*/},
  moredelim=[s][{\itshape\color[rgb]{0,0,0.75}}]{\#[}{]},
  morestring=[b]{"},
  alsodigit={},
  alsoother={},
  alsoletter={!},
  % keywords
  otherkeywords={=>},
  morekeywords={break, continue, else, for, if, in, loop, match, return, while},
  morekeywords={as, const, let, move, mut, ref, static, unsafe},
  morekeywords={dyn, enum, fn, impl, Self, self, struct, trait, type, use, where},
  morekeywords={crate, extern, mod, pub, super},
  morekeywords={abstract, alignof, become, box, do, final, macro,
    offsetof, override, priv, proc, pure, sizeof, typeof, unsized, virtual, yield},
  % traits
  morekeywords=[2]{Send},
  % types
  morekeywords=[3]{bool, char, f32, f64, i8, i16, i32, i64, isize, str, u8, u16, u32, u64, unit, usize, i128, u128},
}%

\def\mytitle{Relational Programming}
\def\myauthor{Yisu Remy Wang}
\def\year{2023}

\title{\mytitle}
\author{\myauthor}

\newcommand\thesisstmt{%
  The future of programming is relational.%
}

\newcommand\Thesisstmt{\expandafter\MakeUppercase \thesisstmt}

\input{macros}

\begin{document}
\pagestyle{empty}

% title and copyright pages
\begin{center}
  {\huge \mytitle}
  \vfill

  {\Large \myauthor}
  \vfill

  \begin{spacing}{1}
    A dissertation \\
    submitted in partial fulfillment of the \\
    requirements for the degree of
  \end{spacing}
  \vfill

  Doctor of Philosophy
  \vfill

  University of Washington \\
  \year
  \vfill

  Reading Commitee: \\
  Dan Suciu, Chair \\
  Zachary Tatlock \\
  Andrew Lumsdaine
  \vfill
  \begin{spacing}{1}
    Program Authorized to Offer Degree: \\
    \cse
  \end{spacing}
  \clearpage

  \textcopyright{} Copyright \year\\
  \myauthor
  \clearpage
\end{center}

\pagestyle{plain}
\setcounter{page}{1}
\pagenumbering{roman}

% abstract page
\begin{center}
  University of Washington \\[1em]
  \textbf{Abstract}        \\[1em]
  \mytitle                 \\[1em]
  \myauthor                \\[1em]

  % Supervisory Committee: \\[-0.5em]
  % Luis Ceze, Chair \qquad
  % Adriana Schulz   \qquad
  % Zachary Tatlock  \\[-0.5em]
  Chair of the Supervisory Committee: \\[-0.5em]
  Dan Suciu \\[-0.5em]
  \pgas
  \\[2em]
\end{center}
\input{abstract}

\begin{spacing}{1.5}
  \tableofcontents
\end{spacing}

\chapter*{Acknowledgments}

By sheer chance I became a student of my advisor Dan Suciu, 
 thus beginning my journey into the world of databases research.
Dan has taught me so much -- not only everything I know about databases, 
 but also to think and speak with clarity, and to ``get to the bottom of it''.
But above all, to be kind. 
Because what matters most is not the facts and figures, 
 but the community of people in pursuit of knowledge.

I'm also proud to call myself a student of Zach Tatlock, 
 even though he's never been my ``advisor on paper''.
When I was lost and confused, Zach welcomed me into the PLSE lab 
 with open arms, as he has done for countless others,
 and supported me unconditionally.
Getting a PhD has been the hardest thing I've ever done, 
 and I knew I could do it because Zach believed in me.

I'm fortunate to call both the PLSE lab and the UWDB lab my home, 
 and I've learned as much from my peers as I have elsewhere.
Among many others, I want to thank Max Willsey, Chandrakana Nandi, 
 Yihong Zhang, Jonathan Leang and Shana Hutchison 
 for sharing this journey with me.

Since I was little, my family have exhausted every last resource 
 to give me the best education possible.
They taught me what it truly means to sacrifice.  
I hope they will be pleased to know that I have taken on 
 a mission to educate the next generation of computer scientists.

When it rains in Seattle, it's always sunny in the small apartment
 I share with my fianc\'{e}e, Ellen Wu.
Thanks for putting up with me,
 and let's go on many more adventures together!

\clearpage

\pagestyle{headings}
\setcounter{page}{1}
\pagenumbering{arabic}
\renewcommand{\chaptermark}[1]{\markboth{\sc{\chaptername\ \thechapter.\ #1}}{}}
\renewcommand{\sectionmark}[1]{\markright{\sc{\thesection.\ #1}}{}}
% \listoftodos
\input{chapters/01-intro}
\input{chapters/related}
\input{chapters/02-background}
\input{chapters/03-datalogo.tex}
\input{chapters/04-free-join.tex}
\input{chapters/05-spores.tex}
\input{chapters/06-fgh.tex}
\input{chapters/07-conclusion.tex}

\clearpage
\addcontentsline{toc}{chapter}{Bibliography}
\singlespacing
\bibliographystyle{alpha}
\bibliography{references}

\end{document}
